Estimating duration using survival analysis

### 2023-08-11 cis-perfect-testing.tex
Restructure first section of perfect testing chapter

\section{Problem description and approach} \label{perf-test:sec:problem}

The CIS infrequently tests each individual, with tests occurring either weekly or monthly (see\todo{refer to section with CIS intro}).
The infrequent testing complicates analysis of the data in two ways: infection episodes can go undetected and the infection episode endpoints (start and end times) are known imprecisely.
This chapter develops a method to account for the infrequent testing pattern.

I approach the problem through survival analysis, the area of statistics concerned with estimating the distribution of the times between two events.
The event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
In the context of this thesis, the distribution to estimate is the duration of positivity.
Therefore, the initiating event is when the duration starts; and the terminating event is when the duration ends.

\todo[inline]{I think the following paragraph is needed somewhere, to answer why I'm not doing the same thing as in ATACCC, but it feels like an aside here since the previous paragraph and the one after this one are more linked.}
Here, I use survival analysis rather than the biomarker approach taken in \autoref{E-ATACCC} for computational efficiency and data limitations.
In the biomarker approach, computational efficiency is an issue because the number of random effects scales with the number of detected infection episodes.
There are a large number of detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
Therefore, the biomarker approach would be computationally challenging with the large sample size in CIS.
The computational challenge is compounded by the data being stored in a trusted research environment which limits the computational power and tools available.
Due to the infrequent sampling in the CIS study design, most detected episodes have only a single positive test, and consequently,
there is no information on the viral load trajectory, and hence the random effects are unlikely to be well-identified.
While the biomarker approach does use more information per sample by taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, the available information is unlikely to be sufficient to overcome these problems.

The first complication in the analysis is that infection episodes can be undetected, meaning no there are no positive tests associated with them.
This happens when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
For these undetected episodes, there is no information on them meaning that we do not even know how many of them occurred.
It is important to account for undetected episodes because they are systematically of shorter duration than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
Undetected infections are a large problem in CIS because the interval between tests is often 28 days (see\todo{ref section on the survey design}), large relative to the duration of an infection episode.
However, we do still know the testing schedule of the individual as they are in the CIS cohort.
Undetected events where we have information on missed events is unusual within the survival analysis literature.
Far more common is that there is no data on these events, a situation known as \emph{truncation}.
A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption[Truncation in CIS data]{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

The second complication is that the endpoints of detected infection episodes are known imprecisely.
An event, such as an endpoint, is known as \emph{censored} if the time of the event can only be bounded.
Censored data are present in almost all survival analyses.
An event is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
%Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
If both the initiating and terminating events are censored, then the event is \emph{doubly censored}.
Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
In CIS data both the initiating and the terminating events are interval censored.
Therefore, the CIS data is \emph{double interval censored}; approaches for double interval censoring are reviewed by \textcite{sunAnalysis,bogaertsSurvival}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

Few studies consider the combination of both double interval censoring and undetected or truncated events
This is especially true within the human biostatistical literature.
While this literature does contain theoretical frameworks which include the double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}, they have been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
The approach of these studies (and elsewhere~\autocite{shenNonparametric}) was to take a conditional likelihood approach, conditioning on both the truncation (a selection effect) and the period in which the initiating event is known to occur.
They justify conditioning on the time of the initiating event by arguing that this information provides little information on the distribution being estimated~\citePersonalComms{Nick Jewell}.
However, this is not the case here.
In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
This phase is not representative of the overall study: infections are more likely to be detected during this phase.
Therefore, estimating the probability of being detected conditional on being infected in this period (which would be implictly done in this previous method)  will overestimate the probability of being detected.

@@

Inference will be performed within the Bayesian paradigm for two reasons.
First, the standard methods for estimating the parameter's variance and covariance are poor.
Commonly used methods are based on the Fisher information matrix or resampling, but neither have theoretical justification~\autocite[230]{sunAnalysis}\todo{Try and understand why this is the case}.
Second, a prior naturally allows the inclusion information from previous studies, such as that from \cref{E-ATACCC}, and/or a belief that the hazard changes smoothly (see \cref{perf-test:sec:parameters-priors}).
These prior structures may be needed to combat the lack of information in the data due to the undetected episodes~\autocite{caoBias}.
They can also be viewed as combining the information from the two analyses (see \cref{perf-test:sec:informative-priors}).

@@

The framework of \textcite{heiseyModelling} formalises this thought experiment.
Start by noting that all relevant information about episode $i$ can be fully characterised by the (unobserved) pair $(b_i, e_i)$ of start and end dates of the episode, which belong to the state space $T \times T$.
If the individual in which the episode occurs is tested between $b_i$ and $e_i$ (inclusive), then the episode is detected, otherwise it is a ghost episode.
For the $i$th detected episode, assume that there are $n_{it}$ ghost episodes.
Further, we discard episodes if there is no negative test prior to their start, and hence also view any such episodes as ghost episodes.

@@

The prior $p(n_i) \propto 1/n_i$ means that the model coincides with the frequentist conditional likelihood approach, up to the priors on other parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}.
Furthermore, this is the reference prior for this quantity~\autocite{heBayesiana}.
Under this prior,
$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})$.
% (\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and
% Rubin} (\protect\hyperlink{ref-dempsterMaximum}{1977}) section 4.2;
% \protect\hyperlink{ref-heiseyModelling}{Heisey and Nordheim}
% (\protect\hyperlink{ref-heiseyModelling}{1995});
% \protect\hyperlink{ref-gelmanBayesian}{Gelman et al.}
% (\protect\hyperlink{ref-gelmanBayesian}{2013}) section 8.7.5)

@@

The form of $S_\theta(t)$ is yet to be specified.
As frequently observed in the literature~\autoref[e.g.:][]{heBayesian}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
This is because the mass function must sum to one and the survival must be monotonically decreasing, while there are no constraints across multiple hazards.
The only constraint on the hazard is to be in the interval $[0, 1]$, since they are probabilities.

@@

I simulate each individual enrolled in CIS who was tested at least once between 10th Oct 2020 and 6th Dec 2020 (inclusive, the same period used for the later analysis using real CIS data).
The individuals' testing schedules are exactly as observed in CIS.
The infection times and durations of positivity are simulated.
From this, a set of test results can be generated.

@@

Individual $i$ is first detectable on their day of infection, $B_i$.
$B_i$ is drawn uniformly between 100 days prior to the first day of testing included and the final day of testing included (inclusive).
$i$ is detectable up to and including the last day of infection, $E_i = B_i + D_i - 1$, where $D_i$ is the duration of positivity for individual $i$, drawn from $f_S$.
Therefore, tests between $B_i$ and $E_i$ (inclusive) are positive, and all other tests are negative.

I then replicate the process used when working with the CIS data by filtering the episodes.
First, discard any episodes where there are no positive tests as they are never detected.
Next, discard any episodes where there are no negative tests prior to the first positive; since there is no lower bound on their infection time, they contain little information, this selection effect was included when constructing $p_{it}$.
Finally, of those remaining episodes, select a random subset to match the number of episodes in the simulation to the number of episodes in the modelling cohort.
The last step is required because everyone is infected in the simulation which does not happen in reality.

For each preserved infection episode, calculate $(l_i^{(b)}, r_i^{b}, l_i^{(e)}, r_i^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.

>>>>

\section{Problem description} \label{perf-test:sec:problem}

The CIS infrequently tests each individual, with tests occurring either weekly or monthly (see\todo{refer to section with CIS intro}).
The infrequent testing complicates analysis of the data in two ways: infection episodes can go undetected and the infection episode endpoints (start and end times) are known imprecisely.
This section explains these issues fully, and the remainder of the chapter develops a method to model them.

The first complication in the analysis is that infection episodes can be undetected, meaning no there are no positive tests associated with them.
This happens when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
For these undetected episodes, there is no information on them episode, meaning that we do not even know how many of them occurred.
It is important to account for undetected episodes because they are systematically of shorter duration than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
Undetected infections are a large problem in CIS because the interval between tests is often 28 days (see\todo{ref section on the survey design}), large relative to the duration of an infection episode.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption[Truncation in CIS data]{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

The second complication is that the beginning and end times of detected infection episodes are known imprecisely.
The imprecision in the beginning of the infection episode arises because the only observation that an episode has started is that an individual was negative at one test, on day $l_i^{(b)}-1$ (the minus one is because the first day that the infection could have begun is the day after the negative test), and positive at the next test, on day $r_i^{(b)}$.
Similarly, the end of the infection episode is bounded by the final positive test on day $l_i^{(e)}$ and the following negative test on day $r_i^{(e)}+1$ (the plus one is because the last day that the infection could have ended is the day before the negative test).
\Cref{perf-test:fig:double-interval-censor} shows this issue graphically.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

I will base my analysis on the 4,800\todo{check number} of episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive and have negatives bounding the start and end time of the episode.
That is, the individual in which the infection episode occurred tested negative at least once in the survey prior to the start of their infection episode, and they tested negative at least once after their infection episode ended.
Without a negative before the episode there is little information on the episode's length because it's start time could have occurred arbitrarily long ago.
The only episodes without a negative following the end of the episode are those in individuals lost to follow-up before their episode ended, which is extremely rare.
This filtering could bias the results of the study but will be considered in the same way as undetected episodes to avoid this.

In the remainder of this and the following chapter, I use survival analysis rather than the biomarker approach taken in \cref{E-ATACCC} for computational efficiency and data limitations arising from the study design.
In the biomarker approach, computational efficiency is an issue because the number of random effects scales with the number of detected infection episodes.
There are a large number of detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
Therefore, the biomarker approach would be computationally challenging with the large sample size in CIS.
The computational challenge is compounded by the data being stored in a TRE with the associated limited access to computational power and software.
The study design of CIS, with the infrequent sampling in the CIS study design, means that most detected episodes have only a single positive test.
Hence, there is no information on the viral load trajectory required to identify the random effects in the model used in \cref{E-ATACCC} or similar approaches.
While the biomarker approach does use more information per sample by taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, the available information is unlikely to be sufficient to overcome these problems.

\section{Survival analysis}

In survival analysis, the event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
In the context of this thesis, the distribution to estimate is the duration of positivity.
Therefore, the initiating event is when the duration starts; and the terminating event is when the duration ends.

Undetected events where we have information on missed events is unusual within the survival analysis literature.
Far more common is that there is no data on these events, a situation known as \emph{truncation}.
A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
A subtlety that makes the CIS different from standard left truncation is that information on individuals in which undetected infections occurred is available because they were enrolled in the study.
This information could be incorporated into the analysis, although doing so increases the computational burden and requires strong assumptions regarding reinfections without noticeably affecting the results (see \cref{appendix:total-model}).
In standard left truncation, nothing is known about the missed events.
As the information is not being incorporated here, standard left truncation methods remain applicable, and they can be modified to incorporate the additional information if desired.

When the time of the initiating and/or terminating events are not known exactly, the survival analysis literature refers to the event(s) as \emph{censored}.
Censored events are present in almost all survival analyses.
An event is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
%Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
If both the initiating and terminating events are censored, then the event is \emph{doubly censored}.
Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
In CIS data both the initiating and the terminating events are interval censored.
Therefore, the CIS data is \emph{double interval censored}; approaches for double interval censoring are reviewed by \textcite{sunAnalysis,bogaertsSurvival}.

Few studies consider the combination of both double interval censoring and undetected or truncated events
This is especially true within the human biostatistical literature.
While this literature does contain theoretical frameworks which include the double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}, they have been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
The approach of these studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} was to take a conditional likelihood approach, conditioning on both the truncation (a selection effect) and the period in which the initiating event is known to occur.
They justify conditioning on the time of the initiating event by arguing that this information provides little information on the distribution being estimated~\citePersonalComms{Nick Jewell}.
However, this is not the case here.
In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
This phase is not representative of the overall study: infections are more likely to be detected during this phase.
Therefore, estimating the probability of being detected conditional on being infected in this period (which would be implictly done in this previous method)  will overestimate the probability of being detected.

@@

Inference will be performed within the Bayesian paradigm for two reasons.
First, the standard frequentist methods for estimating the parameter's variance and covariance are poorly developed in this area.
Commonly used methods are based on the Fisher information matrix or resampling, but neither have theoretical justification~\autocite[230]{sunAnalysis}\todo{Try and understand why this is the case}.
Second, a prior naturally allows the inclusion information from previous studies, such as that from \cref{E-ATACCC}, and/or a belief that the hazard changes smoothly in a way that can be encoded in a prior (these options are explored in \cref{perf-test:sec:parameters-priors}).
These prior structures may be needed to combat the lack of information in the data due to the undetected episodes~\autocite{caoBias}.
They can also be viewed as combining the information from the two analyses (see \cref{perf-test:sec:informative-priors}).

@@

The framework of \textcite{heiseyModelling} formalises this thought experiment.
Start by noting that all relevant information about episode $i$ can be fully characterised by the (unobserved) pair $(b_i, e_i)$ of start and end dates of the episode, which belong to the state space $T \times T$.
If the individual in which the episode occurs is tested between $b_i$ and $e_i$ (inclusive), then the episode is detected, otherwise it is a ghost episode.
For the $i$th detected episode, assume that there are $n_{it}$ ghost episodes.

@@

The reference prior for $n_i$ is $p(n_i) \propto 1/n_i$~\autocite{heBayesiana}.
Under this prior,
$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})$.
This likelihood is equal to the conditional likelihood used in the frequentist approaches previously mentioned.
It has been used elsewhere for Bayesian inference~\autocite{zhouUnified}, but the justification within a Bayesian paradigm has not been made explicitly.
%The prior $p(n_i) \propto 1/n_i$ means that the model coincides with the frequentist conditional likelihood approach, up to the priors on other parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}.
% (\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and
% Rubin} (\protect\hyperlink{ref-dempsterMaximum}{1977}) section 4.2;
% \protect\hyperlink{ref-heiseyModelling}{Heisey and Nordheim}
% (\protect\hyperlink{ref-heiseyModelling}{1995});
% \protect\hyperlink{ref-gelmanBayesian}{Gelman et al.}
% (\protect\hyperlink{ref-gelmanBayesian}{2013}) section 8.7.5)

@@

The form of $S_\theta(t)$ is yet to be specified.
As frequently observed in the literature~\autocite[e.g.:][]{heBayesian}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
This is because the mass function must sum to one and the survival must be monotonically decreasing, while there are no constraints across multiple hazards.
The only constraint on the hazard is to be in the interval $[0, 1]$, since they are probabilities.

@@

I simulate a dataset of detected episodes that has the same characteristics as that in the CIS by the following procedure.
\begin{enumerate}
    \item Extract the test schedules for each individual who had at least one test during the period of interest.
    \item Draw an episode start time, $b_i$ for each individual uniformly at random between 2nd July 2020 (100 days before the period where a detected episode would be included) and 6th December 2020 (the end of this period).
    \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_i = b_i + d_i - 1$.
    \item Simulate the test results based on the test schedule, $b_i$, and $e_i$. Tests between $b_i$ and $e_i$ (inclusive) are positive, all other tests are negative.
    \item Discard episodes where there are no positive tests (\ie undetected episodes) and then apply the same inclusions/exclusions as in \cref{perf-test:sec:problem}.
    \item Of these remaining episodes, sample 4,800 to match the sample size of the true dataset. This is needed because in step 2 the entire cohort was infected, while in the real study only a (unknown) portion is infected.
    \item For this final set of episodes, calculate $(l_i^{(b)}, r_i^{b}, l_i^{(e)}, r_i^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
\end{enumerate}

@@
